---
title: "Interactive Edge Editor (Minimal Shiny App)"
author: "Anthony Wu"
package: MSstatsBioNet
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{MSstatsBioNet: Interactive Edge Editor (Minimal Shiny App)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>"
)
```

# Overview

This vignette shows a self-contained Shiny app that lets you:

1. Supply **nodes** and **edges** — either as built-in example data or by uploading
   your own CSV files.
2. **Visualise** the network with `cytoscapeNetwork()`.
3. **Delete edges** interactively by Ctrl+clicking or right-clicking any edge.
4. **Download analysis code** — an R script that reproduces the curated network,
   including a `deleteEdgeFromNetwork()` call for every edge you removed.

The app requires only `shiny` and `MSstatsBioNet`; no INDRA API calls or MSstats
pipeline are involved.

# CSV format

If you choose to upload your own data the expected column layout is:

| File | Required columns | Optional columns |
|------|-----------------|-----------------|
| Nodes CSV | `id` | `logFC`, `hgncName`, `Site` |
| Edges CSV | `source`, `target`, `interaction` | `site`, `evidenceLink` |

# App code

Copy the block below into an `.R` file and call `shinyApp(ui, server)` to launch,
or uncomment the last line and source the file directly.

```{r eval = FALSE}
if (!requireNamespace("shiny", quietly = TRUE)) {
    stop("Install the 'shiny' package to run this app.")
}

library(shiny)
library(MSstatsBioNet)

# ── Built-in example data ─────────────────────────────────────────────────────

example_nodes <- data.frame(
    id    = c("TP53", "MDM2", "CDKN1A", "BCL2", "BAX", "CASP3"),
    logFC = c( 1.8,   -0.9,    2.1,    -1.5,    1.2,   0.7),
    stringsAsFactors = FALSE
)

example_edges <- data.frame(
    source      = c("TP53",       "TP53",           "MDM2",
                    "TP53",       "BCL2",            "BAX"),
    target      = c("MDM2",       "CDKN1A",         "TP53",
                    "BAX",        "BAX",             "CASP3"),
    interaction = c("Activation", "IncreaseAmount",  "Inhibition",
                    "Activation", "Inhibition",      "Activation"),
    stringsAsFactors = FALSE
)

# ── UI ────────────────────────────────────────────────────────────────────────

ui <- fluidPage(
    titlePanel("Network Edge Editor"),

    sidebarLayout(
        sidebarPanel(
            width = 3,

            # Data source
            radioButtons(
                "data_source", "Data source",
                choices  = c("Example data" = "example",
                             "Upload CSV files" = "upload"),
                selected = "example"
            ),
            conditionalPanel(
                condition = "input.data_source == 'upload'",
                fileInput("nodes_file", "Nodes CSV", accept = ".csv"),
                tags$small(
                    "Required column: ", tags$code("id"), ". ",
                    "Optional: ", tags$code("logFC"), ", ",
                    tags$code("hgncName"), "."
                ),
                br(),
                fileInput("edges_file", "Edges CSV", accept = ".csv"),
                tags$small(
                    "Required columns: ", tags$code("source"), ", ",
                    tags$code("target"), ", ", tags$code("interaction"), "."
                )
            ),

            hr(),
            actionButton(
                "show_network", "Display Network",
                class = "btn-primary", width = "100%"
            ),

            hr(),
            tags$p(
                tags$strong("To delete an edge:"),
                " Ctrl+click or right-click it in the network."
            ),

            # Live deletion log
            uiOutput("deleted_summary"),

            hr(),
            # Download button appears once network has been rendered
            uiOutput("download_ui")
        ),

        mainPanel(
            width = 9,
            cytoscapeNetworkOutput("network", height = "600px")
        )
    )
)

# ── Server ────────────────────────────────────────────────────────────────────

server <- function(input, output, session) {

    # Accumulates edge info lists: list(source, target, interaction)
    deleted_edges <- reactiveVal(list())

    # ── Resolve input data ────────────────────────────────────────────────────

    nodes_data <- reactive({
        if (input$data_source == "example") return(example_nodes)
        req(input$nodes_file)
        read.csv(input$nodes_file$datapath, stringsAsFactors = FALSE)
    })

    edges_data <- reactive({
        if (input$data_source == "example") return(example_edges)
        req(input$edges_file)
        read.csv(input$edges_file$datapath, stringsAsFactors = FALSE)
    })

    # ── Render network ────────────────────────────────────────────────────────

    observeEvent(input$show_network, {
        # Fresh render → reset the deletion log
        deleted_edges(list())

        output$network <- renderCytoscapeNetwork({
            cytoscapeNetwork(nodes = nodes_data(), edges = edges_data())
        })

        # Reveal download button
        output$download_ui <- renderUI({
            downloadButton("download_code", "Download analysis code")
        })
    })

    # ── Track deletions ───────────────────────────────────────────────────────

    # The cytoscapeNetwork JS widget fires input$network_edge_deleted
    # (with fields $source, $target, $interaction) on Ctrl+click / right-click.
    observeEvent(input$network_edge_deleted, {
        e <- input$network_edge_deleted
        deleted_edges(c(deleted_edges(), list(e)))
    })

    # Live summary shown in the sidebar
    output$deleted_summary <- renderUI({
        deleted <- deleted_edges()
        if (length(deleted) == 0) return(NULL)
        items <- lapply(deleted, function(e) {
            tags$li(paste0(e$source, " \u2192 ", e$target,
                           " (", e$interaction, ")"))
        })
        tagList(
            tags$p(tags$strong(paste(length(deleted), "edge(s) deleted:"))),
            tags$ul(items)
        )
    })

    # ── Download analysis code ────────────────────────────────────────────────

    output$download_code <- downloadHandler(
        filename = function() {
            paste0("network-analysis-", Sys.Date(), ".R")
        },
        content = function(file) {
            # Capture the data frames as self-contained R expressions
            nodes_expr <- paste(
                capture.output(dput(isolate(nodes_data()))),
                collapse = "\n"
            )
            edges_expr <- paste(
                capture.output(dput(isolate(edges_data()))),
                collapse = "\n"
            )

            lines <- c(
                "library(MSstatsBioNet)",
                "",
                "# Nodes",
                paste0("nodes <- ", nodes_expr),
                "",
                "# Edges",
                paste0("edges <- ", edges_expr),
                ""
            )

            deleted <- isolate(deleted_edges())
            if (length(deleted) > 0) {
                lines <- c(lines, "# Delete edges removed interactively")
                for (e in deleted) {
                    lines <- c(lines, sprintf(
                        'edges <- deleteEdgeFromNetwork(edges, "%s", "%s", "%s")',
                        e$source, e$target, e$interaction
                    ))
                }
                lines <- c(lines, "")
            }

            lines <- c(
                lines,
                "# Visualise network",
                "cytoscapeNetwork(nodes, edges)",
                "",
                "# Export to a standalone HTML file",
                "exportNetworkToHTML(nodes, edges)"
            )

            writeLines(lines, file)
        }
    )
}

# shinyApp(ui, server)   # uncomment to launch
```

# What the downloaded script looks like

After displaying the network and deleting, say, the `MDM2 → TP53 (Inhibition)` and
`BCL2 → BAX (Inhibition)` edges, clicking **Download analysis code** produces an `.R`
file like this:

```r
library(MSstatsBioNet)

# Nodes
nodes <- structure(list(...), class = "data.frame")

# Edges
edges <- structure(list(...), class = "data.frame")

# Delete edges removed interactively
edges <- deleteEdgeFromNetwork(edges, "MDM2", "TP53", "Inhibition")
edges <- deleteEdgeFromNetwork(edges, "BCL2", "BAX", "Inhibition")

# Visualise network
cytoscapeNetwork(nodes, edges)

# Export to a standalone HTML file
exportNetworkToHTML(nodes, edges)
```

The script is fully self-contained: running it from top to bottom reproduces the
exact curated network without needing the Shiny app.

# Session info

```{r}
sessionInfo()
```
