---
title: "PTM Analysis"
author: "Anthony Wu"
package: MSstatsBioNet
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{MSstatsBioNet: PTM Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


# Installation

Run this code below to install MSstatsBioNet from bioconductor

```{r eval = FALSE}
# if (!require("BiocManager", quietly = TRUE)) {
#     install.packages("BiocManager")
# }
# 
# BiocManager::install("MSstatsBioNet")

install.packages("devtools")
devtools::install_github("Vitek-Lab/MSstatsBioNet@feature-ptm", build_vignettes = TRUE)
```

## Dataset

We will be taking a subset of the dataset found in this [paper](https://www.biorxiv.org/content/10.1101/2024.10.21.619348v1).  The table is the output of the MSstatsPTM function `groupComparisonPTM` (filtered down to the columns that are actually needed)

```{r}
input = data.table::fread(system.file(
    "extdata/2_egf_data.csv",
    package = "MSstatsBioNet"
))
head(input)
```

I'm going to do some renaming of the columns such that the PTM analysis goes smoothly.  Specifically, the `GlobalProtein` column will be renamed to `Protein` and the `Protein` column will be renamed to `Site`.  In the future, you won't need to do this.
```{r}
library(tidyverse)
input <- input %>%
    rename(Site = Protein, Protein = GlobalProtein)
```

## ID Conversion

First, we need to convert the group comparison results to a format that can be
processed by INDRA.  We can use the `annotateProteinInfoFromIndra` function
to obtain these mappings.

In the below example, we convert uniprot IDs to their corresponding Hgnc IDs. We
can also extract other information, such as hgnc gene name and protein function.

```{r}
library(MSstatsBioNet)
annotated_df = annotateProteinInfoFromIndra(input, "Uniprot")
head(annotated_df)
```


## getPathwaysFromIndra

Here, we will get pathways from INDRA and rank them based on relevance.  We will see here that GAB1 is rated as the most relevant edge to track.

```{r}
pathways = getPathwaysFromIndra(annotated_df, main_target = "P00533")
# optional: filter by phosphorylation
pathways$edges = pathways$edges %>% filter(str_detect(interaction, "Phosphorylation"))
```

## Visualize Networks

The function `visualizeNetworks` then takes the output of 
`getSubnetworkFromIndra` and visualizes the subnetwork. The function requires
Cytoscape desktop to be open for the visualization to work.

```{r}
visualizeNetworks(subnetwork$nodes, subnetwork$edges)
```

## Subnetwork Query

The package provides a function `getSubnetworkFromIndra` that retrieves a
subnetwork of proteins from the INDRA database based on differential abundance
analysis results.  This function may help finding off target subnetworks.  

```{r}
subnetwork <- getSubnetworkFromIndra(annotated_df, pvalueCutoff = 0.05)
head(subnetwork$nodes)
head(subnetwork$edges)
```

# Session info

```{r}
sessionInfo()
```
